\documentclass{beamer}

\input{preamble.tex}

\title{OpenMP for Computational Scientists}
\subtitle{Preliminaries}

\begin{document}

\frame{\titlepage}

%-------------------------------------------------------------------------------
\begin{frame}
\frametitle{Audience}

\begin{itemize}
\item Teaches OpenMP 4.5 and 5.0 in a seminar style.
\item 6 lecture topics, with exercises and solutions.
\item Designed for Computational Scientists familiar with Fortran and MPI programming.

\end{itemize}

Download code (and slides) from:
\url{https://github.com/UoB-HPC/openmp-for-cs}

\end{frame}
%-------------------------------------------------------------------------------

\begin{frame}
\frametitle{Exercises}
\begin{itemize}
\item This is a hands-on course!
\item Exercises will be set for you to try programming OpenMP yourselves.
\item Sample solutions also provided.
\item All the exercises will be in Fortran.
\end{itemize}

\end{frame}

%-------------------------------------------------------------------------------
\section{Outline}
\begin{frame}
\frametitle{Course Outline}
Organised as 6 sessions teaching OpenMP plus top-tips for getting good performance.
\begin{enumerate}
  \item OpenMP overview
  \item Data sharing and reductions
  \item Vectorisation and code optimisations
  \item NUMA and MPI interoperability
  \item GPU programming with OpenMP
  \item Tasks and Tools
\end{enumerate}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
\frametitle{Thanks}
Thanks go to the following authors, whose own OpenMP tutorials have inspired this one:
\begin{itemize}
  \item Tim Mattson (Intel)
  \item Alice Koniges (Berkeley Lab/NERSC)
  \item Simon McIntosh-Smith and the HPC team (UoBristol)
  \item Gethin Williams (UoBristol)
  \item and many others
\end{itemize}
\end{frame}
%-------------------------------------------------------------------------------

\end{document}
