\documentclass[aspectratio=169]{beamer}

\input{preamble.tex}

\title{OpenMP for Computational Scientists}
\subtitle{Wrap up}

\begin{document}

\frame{\titlepage}

%-------------------------------------------------------------------------------

\begin{frame}
\frametitle{OpenMP 5.0}
OpenMP 5 adds features to make writing performance portable programs simpler.

Highlighting some applicable to target:
\begin{itemize}
\item Loop construct
\item Mappers
\item Unified Shared Memory (USM)
\item Function variants
\item Reverse offload
\item \mintinline{fortran}|OMP_TARGET_OFFLOAD|
\item Reduction variables now implicitly \mintinline{fortran}|map(tofrom)|
\end{itemize}

\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}[fragile]
\frametitle{Loop}
\begin{itemize}
\item Assert that the iterations in a loop nest may execute in any order, including concurrently
\item Let the compiler figure our how to best utilize parallel resources
\end{itemize}

\begin{minted}[]{fortran}
!$omp target
!$omp loop
do i = 1, N
  a(i) = b(i)
end do
!$omp end loop
!$omp end target
\end{minted}

\end{frame}
%-------------------------------------------------------------------------------

\begin{frame}[fragile]
\frametitle{Unified shared memory}
Code requires specific features, e.g. shared memory between host and devices.

\begin{minted}[]{fortran}

real(kind=8), dimension(:), allocatable :: A
allocate(A(1024))

!$omp requires unified_shared_memory

!$omp target
  call do_something_with_A(A)
!$omp end target
\end{minted}

No map clauses. Data is shared between the host and device.

\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
\frametitle{OpenMP resources}
\begin{itemize}
\item Two brilliant books from MIT Press:
\begin{itemize}
\item The OpenMP Common Core: Making OpenMP Simple Again --- Tim Mattson, Yun (Helen) Ye and Alice Koniges.
\item Using OpenMP - The Next Steps --- Ruud van de Pas, Eric Stotzer and Christian Terboven.
\end{itemize}
 \item OpenMP website: \url{https://www.openmp.org}
   \begin{itemize}
     \item The specification (not for the faint hearted).
     \item Download summary cards.
     \item List of compiler support.
     \item Example code for all the directives.
     \item List of books: \url{https://www.openmp.org/resources/openmp-books/}
   \end{itemize}
 
\end{itemize}
\end{frame}
%-------------------------------------------------------------------------------
\end{document}
